Empirical normative theories

### 2020-07-18 simple.tex
Slightly larger obligations for empirical theory

\section{Additive single-answer empirical theories}
\subsection{Definition}
\commentOC{I leave the definition as is, but for the moment I only consider ${\gleadsto}$, a relation that does not depend on $i$; and I do not use $q(i)$.}
An empirical theory for a normative theory is a tuple $(q, (\gleadsto[q(i)])_{q(i) \in q(I)}, \gbeats, \alldargs, \dbeats, +)$,
with ${\gleadsto[q(i)]} \subseteq (\allargs × \Phi) \cap (\alldargs × \Phi)$, 
${\gbeats} \subseteq \allargs × \allargs$ and is irreflexive,
${\dbeats} \subseteq \alldargs × \allargs$ and is injective and its range equals $\gbeatsinv(\allargs)$, the domain of ${\gbeats}$ (thus $\forall \ar_1 \suchthat [\exists \ar_0 \suchthat \ar_1 \gbeats \ar_0]$: $\dbeatsinv(\ar_1)$ is a singleton). 
Define $\df \subseteq \allargs × \alldargs$ as $\df = \dbeatsinv \circ \gbeatsinv$. 
The symbol $+$ in an empirical theory stands for a function $+: \bigcup_{\ar \in \allargs}(\set{s} × \df(\ar)) → \allargs$.
As usual, $\ar + \ard$ means $+(\ar, \ard)$. 
Define $\dfp \subseteq \allargs × \allargs$ as, $\forall \ar \in \allargs$: $\dfp(\ar) = \ar + \df(\ar) = \set{\ar + \ard \suchthat \ard \in \df(\ar)}$.
An empirical theory must satisfy
$\exists k \in \N \suchthat \dfp^k(\gleadstoinv(\allprops)) = \emptyset$.

@@

Given $i \in I, \phi \in \gPhi, \ar \in \allargs$, define the \emph{$\phi$-defensors of $\ar$ required for $i$}, $\di(\ar) \subseteq \alldargs$, as $\di(\ar) = [\dbeatsinv \circ {\ibeatsinv}](\ar)$. Define $\dip: \allargs → \allargs$ as $\dip(\ar) = \ar + \di(\ar)$.
Given $i \in I, \phi \in \gPhi$, define $\gargs[i, \phi] \subseteq \allargs$ as the transitive closure of ${\gleadstoinv}(\phi)$ under $\dip$: $\gargs[i, \phi] = \cup_{k \in \N} (\dip)^k({\gleadstoinv}(\phi))$, where $(\dip)^0$ is the identity function over $\allargs$.

@@

\subsection{Effectively comparing models}
Model: for a given $i$. Defends $\prop$. $\exists \ar_1 \in \args_\eta \suchthat \ar_1 \ileadsto \prop$. Then, $\forall \ar_2 \in \gargs \suchthat (\ar_2, \ar_1) \ileadsto ¬\prop: \exists \ar_3 \in \args_\eta \suchthat (\ar_3, \ar_2) \ileadsto \prop.$

@@

\section{Decision situation}
\NewDocumentCommand{\ileadstoprop}{}{⇝^\prop}
\NewDocumentCommand{\ileadstonprop}{}{⇝^{¬\prop}}
\NewDocumentCommand{\ileadstosts}{}{⇝_*}
\NewDocumentCommand{\ileadstos}{}{\overset{s}{⇝_i}}
\NewDocumentCommand{\ileadstostw}{O{}}{⇝^\mathit{w}_\forall}
\NewDocumentCommand{\ileadstostp}{O{}}{⇝^\mathit{pos}_\forall}
\NewDocumentCommand{\nileadstosts}{}{\not⇝_*}
\NewDocumentCommand{\Ti}{}{T_i}
\NewDocumentCommand{\ibeatse}{O{}}{⊳^{#1}_\exists}
\NewDocumentCommand \ibeatseinv { o }{
	\IfValueTF{#1}{%
		{⊳_\exists^{#1}}^{-1}%
	}{%
		⊳_\exists^{-1}%
	}%
}
\NewDocumentCommand{\zar}{}{\mathbf{0}}%zero, or empty, argument
\NewDocumentCommand{\mPhi}{}{\Phi_\gamma}
\NewDocumentCommand{\argsd}{}{S^\mathit{d}}
\RenewDocumentCommand{\ileadstost}{O{}}{⇝^{#1}_\forall}
\RenewDocumentCommand{\nileadstost}{O{}}{\not⇝^{#1}_\forall}

The object of study in a given decision situation is the deliberated preferences of a given individual $i$. 
A decision situation concerns a topic $\allprops$, given a set of arguments $\allargs$, given a certain query protocol, and over a certain time frame. The relation $\ileadstost$, to be defined shortly, models the reaction of $i$ to arguments about the topic. The decision situation will be defined as a triple $(\allprops, \allargs, \ileadstost)$ satisfying some conditions to be defined after having presented those three fundamental elements.

@@

Define $\Phi = \allprops × \set{\text{possible}, \text{sure}}$. Elements in $\Phi$ will also be called propositions, which should create no ambiguity. 
The relation $\ileadstost \subseteq (\allargs × \allargs) × \Phi$, pronounced “always leads to”, has the following semantics.
Whenever $(\ar, \ar') \ileadstost (\prop, \text{sure})$, $i$ always considers that $\prop$ is sure (meaning that $¬\prop$ is excluded), when presented with these two arguments, and whenever $(\ar, \ar') \ileadstost (\prop, \text{possible})$, $i$ always considers that $\prop$ is “at least” possible (meaning that $¬\prop$ may be possible as well), when presented with these two arguments. In both cases, the term “always” means that the consideration is stable over time, including after having presented other arguments to $i$. 
Two query protocols presented below will illustrate how these semantics might be satisfied. 
However, I voluntarily leave the querying protocol not precisely specified in general, as multiple reasonable choices are possible. The results of this article depend only on assumptions about $\ileadstost$.

Define the negation operator $¬$ over $\Phi$ as: $¬(\prop, \text{possible}) = (¬\prop, \text{sure})$ and $¬(\prop, \text{sure}) = (¬\prop, \text{possible})$. It follows that $¬¬\phi = \phi$. Write $(\ar, \ar') \nileadstost \phi$ for $¬[(\ar, \ar') \ileadstost \phi]$ and define $(\ar, \ar') \ileadstoe \phi$ as equivalent to $(\ar, \ar') \nileadstost ¬\phi$. 
The relation $\ileadstoe$ is pronounced “sometimes leads to”.
A proposition $\phi$ is about $\prop$ iff $\phi \in \{\prop, ¬\prop\} × \{\text{possible}, \text{sure}\}$.

The pair $(\ar, \ar')$ is to be understood as unordered (thus $(\ar, \ar') \ileadstost \prop ⇔ (\ar', \ar) \ileadstost \prop$).

\begin{definition}[Decision situation]
	A decision situation is a topic $\allprops$ closed under negation, a set of arguments $\allargs$, an “always leads to” relation $\ileadstost \subseteq (\allargs × \allargs) × \Phi$ satisfying (A1) and (A2) and using unordered pairs of arguments, $\Phi$ being the set of propositions $\allprops × \set{\text{sure}, \text{possible}}$.
\end{definition}

\subsection{Query protocols}
As vNM aptly summarize, “It is clear that every measurement – or rather every claim of measurability – must ultimately be based on some immediate sensation, which possibly cannot and certainly need not be analyzed any futher.” The fundamental element here is the reaction of $i$ to arguments, which will serve as the informational basis to define $\ileadstost$. Given a proposition $\prop$ and two arguments $\ar, \ar' \in \allargs$, a query consists in presenting both arguments to $i$ and observing which proposition $i$ considers valid in her current state of mind, thus using both arguments and possibly other arguments she has in mind: is it that $\prop$ holds, that $¬\prop$ holds, or that both are possible? The third possibility allows for the case where no argument appears more decisive than the other one.

@@

Define $\ileadstost$ from $\ileadstoprop$ as follows, for any $\prop \in \allprops$: $(\ar, \ar') \ileadstost (\prop, \text{possible}) ⇔ (\ar, \ar') \ileadstoprop C$ with $C \subseteq \set{\prop, \text{both}}$; and $(\ar, \ar') \ileadstost (\prop, \text{sure}) ⇔ (\ar, \ar') \ileadstoprop \set{\prop}$.

@@

When “both” is not observable, $\ileadstost$ is defined as: $(\ar, \ar') \ileadstost (\prop, \text{possible}) ⇔ (\ar, \ar') \ileadstost (\prop, \text{sure}) ⇔ (\ar, \ar') \ileadstoprop \set{\prop}$. (A1) and (A2) still hold. We say in this case that $\ileadstost$ does not distinguish sure from possible. Thus, elements $\phi \in \Phi$ may be considered as simply equal to $\prop$ or $¬\prop$, as $(\prop, \text{sure})$ is treated exactly as $(\prop, \text{possible})$.

@@

\begin{remark}
	In fact, the model is tasked with finding a stable point after interrogating $i$. The process of trying counter-arguments $\ar_1, …$ is $q(i)$. When done, the model claims that $\ar_0 + \ar_2 + …$ is decisive. The model also claims that another process of interrogation does not lead to different answers. This is falsifiable.
	When claiming that $(\ar_1, \ar_0) \ileadstost[i] \phi$, the model really claims that $\nexists p, i' \in q^{-1}(q(i)) \suchthat p, \ar_1, \ar_0 \ileadsto[i'] ¬\phi \lor p, \ar_0, \ar_1 \ileadsto[i'] ¬\phi$, where $i' \in q^{-1}(q(i))$ iff $q(i) = q(i')$.

>>>>

\section{Additive single-answer empirical theories}
\subsection{Definition}
\commentOC{I leave the definition as is, but for the moment I only consider ${\gleadsto}$, a relation that does not depend on $i$; and I do not use $q(i)$.}
An empirical theory for a normative theory is a tuple $(q, (\gleadsto[q(i)])_{q(i) \in q(I)}, \gbeats, \alldargs, \dbeats, +)$,
with ${\gleadsto[q(i)]} \subseteq (\allargs × \Phi) \cap (\alldargs × \Phi)$, 
${\gbeats} \subseteq \allargs × \allargs$ and is irreflexive,
${\dbeats} \subseteq \alldargs × \allargs$ and is injective and its range includes $\gbeatsinv(\allargs)$, the domain of ${\gbeats}$ (thus $\forall \ar_1 \suchthat [\exists \ar_0 \suchthat \ar_1 \gbeats \ar_0]$: $\dbeatsinv(\ar_1)$ is a singleton). 
Define $\df \subseteq \allargs × \alldargs$ as $\df = \dbeatsinv \circ \gbeatsinv$. 
The symbol $+$ in an empirical theory stands for a function $+: \bigcup_{\ar \in \allargs}(\set{s} × \df(\ar)) → \allargs$.
As usual, $\ar + \ard$ means $+(\ar, \ard)$. 
Define $\dfp \subseteq \allargs × \allargs$ as, $\forall \ar \in \allargs$: $\dfp(\ar) = \ar + \df(\ar) = \set{\ar + \ard \suchthat \ard \in \df(\ar)}$.
An empirical theory must satisfy
$\exists k \in \N \suchthat \dfp^k(\dfp(\zar) \cup \gleadstoinv(\Phi)) = \emptyset$, where $\dfp^0$ is the identity function over $\allargs$.

@@

Given $i \in I, \phi \in \gPhi, \ar \in \allargs$, define the \emph{$\phi$-defensors of $\ar$ required for $i$}, $\di(\ar) \subseteq \alldargs$, as $\di(\ar) = [\dbeatsinv \circ {\ibeatsinv}](\ar)$. Define $\dip: \allargs → \allargs$ as $\dip(\ar) = \ar + \di(\ar)$.
Given $i \in I, \phi \in \gPhi$, define $\gargs[i, \phi] \subseteq \allargs$ as the transitive closure of ${\gleadstoinv}(\phi)$ under $\dip$: $\gargs[i, \phi] = \cup_{k \in \N} (\dip)^k(\dip(\zar) \cup {\gleadstoinv}(\phi))$.

@@

\subsection{Effectively comparing models}
Given a normative theory, consider two empirical theories $(q, (\gleadsto[q(i)])_{q(i) \in q(I)}, \gbeats, \alldargs, \dbeats, +)$ and $(q', (\gleadsto[q(i)][\delta])_{q(i) \in q(I)}, \gbeats', \alldargs, \dbeats', +')$.

Given a normative theory and $\phi \in \Phi$, say that two empirical theories (for that normative theory) disagree about $\phi$ iff $\phi \in \gPhi \land ¬\phi \in \gPhi[\delta]$.

If $\gamma$ and $\delta$ agree about all $\phi$, there is nothing to test as the theories are compatible. Otherwise, here is a test procedure for that falsifies at least one of these theories, given any $\phi \in \Phi$ about which they disagree and any $i \in I$.

Note first that $\ar_0 \gbeats \ar_0$ is impossible because … TODO why didn’t I need this before?

Consider $\ar_0 \in \allargs \suchthat \ar_0 \gleadsto \phi$. Test whether $(\ar_0) \ileadstost \phi$. If not, $(i, \phi, (\ar_0))$ is a falsification instance of $\gamma$. If $(\ar_0) \ileadsto \phi$, check whether some $\ar_1 \dbeats' \ar_0$. If not, $\delta$ is falsified? ($\ar_0$ attacks $\delta$ but $\delta$ is not able to defend.)…

Model: for a given $i$. Defends $\prop$. $\exists \ar_1 \in \args_\eta \suchthat \ar_1 \ileadsto \prop$. Then, $\forall \ar_2 \in \gargs \suchthat (\ar_2, \ar_1) \ileadsto ¬\prop: \exists \ar_3 \in \args_\eta \suchthat (\ar_3, \ar_2) \ileadsto \prop.$

@@

\section{Decision situation}
\NewDocumentCommand{\ileadstoprop}{}{⇝^\prop}
\NewDocumentCommand{\ileadstonprop}{}{⇝^{¬\prop}}
\NewDocumentCommand{\ileadstosts}{}{⇝_*}
\NewDocumentCommand{\ileadstos}{}{\overset{s}{⇝_i}}
\NewDocumentCommand{\ileadstostw}{O{}}{⇝^\mathit{w}_\forall}
\NewDocumentCommand{\ileadstostp}{O{}}{⇝^\mathit{pos}_\forall}
\NewDocumentCommand{\nileadstosts}{}{\not⇝_*}
\NewDocumentCommand{\Ti}{}{T_i}
\NewDocumentCommand{\ibeatse}{O{}}{⊳^{#1}_\exists}
\NewDocumentCommand \ibeatseinv { o }{
	\IfValueTF{#1}{%
		{⊳_\exists^{#1}}^{-1}%
	}{%
		⊳_\exists^{-1}%
	}%
}
\NewDocumentCommand{\mPhi}{}{\Phi_\gamma}
\NewDocumentCommand{\argsd}{}{S^\mathit{d}}
\NewDocumentCommand{\ileadstoall}{O{}}{⇝^{#1}_\forall}
\NewDocumentCommand{\nileadstoall}{O{}}{\not⇝^{#1}_\forall}

The object of study in a given decision situation is the deliberated preferences of a given individual $i$. 
A decision situation concerns a topic $\allprops$, given a set of arguments $\allargs$, given a certain query protocol, and over a certain time frame. The relation $\ileadstoall$, to be defined shortly, models the reaction of $i$ to arguments about the topic. The decision situation will be defined as a triple $(\allprops, \allargs, \ileadstoall)$ satisfying some conditions to be defined after having presented those three fundamental elements.

@@

Define $\Phi = \allprops × \set{\text{possible}, \text{sure}}$. Elements in $\Phi$ will also be called propositions, which should create no ambiguity. 
The relation $\ileadstoall \subseteq (\allargs × \allargs) × \Phi$, pronounced “always leads to”, has the following semantics.
Whenever $(\ar, \ar') \ileadstoall (\prop, \text{sure})$, $i$ always considers that $\prop$ is sure (meaning that $¬\prop$ is excluded), when presented with these two arguments, and whenever $(\ar, \ar') \ileadstoall (\prop, \text{possible})$, $i$ always considers that $\prop$ is “at least” possible (meaning that $¬\prop$ may be possible as well), when presented with these two arguments. In both cases, the term “always” means that the consideration is stable over time, including after having presented other arguments to $i$. 
Two query protocols presented below will illustrate how these semantics might be satisfied. 
However, I voluntarily leave the querying protocol not precisely specified in general, as multiple reasonable choices are possible. The results of this article depend only on assumptions about $\ileadstoall$.

Define the negation operator $¬$ over $\Phi$ as: $¬(\prop, \text{possible}) = (¬\prop, \text{sure})$ and $¬(\prop, \text{sure}) = (¬\prop, \text{possible})$. It follows that $¬¬\phi = \phi$. Write $(\ar, \ar') \nileadstoall \phi$ for $¬[(\ar, \ar') \ileadstoall \phi]$ and define $(\ar, \ar') \ileadstoe \phi$ as equivalent to $(\ar, \ar') \nileadstoall ¬\phi$. 
The relation $\ileadstoe$ is pronounced “sometimes leads to”.
A proposition $\phi$ is about $\prop$ iff $\phi \in \{\prop, ¬\prop\} × \{\text{possible}, \text{sure}\}$.

The pair $(\ar, \ar')$ is to be understood as unordered (thus $(\ar, \ar') \ileadstoall \prop ⇔ (\ar', \ar) \ileadstoall \prop$).

\begin{definition}[Decision situation]
	A decision situation is a topic $\allprops$ closed under negation, a set of arguments $\allargs$, an “always leads to” relation $\ileadstoall \subseteq (\allargs × \allargs) × \Phi$ satisfying (A1) and (A2) and using unordered pairs of arguments, $\Phi$ being the set of propositions $\allprops × \set{\text{sure}, \text{possible}}$.
\end{definition}

\subsection{Query protocols}
As vNM aptly summarize, “It is clear that every measurement – or rather every claim of measurability – must ultimately be based on some immediate sensation, which possibly cannot and certainly need not be analyzed any futher.” The fundamental element here is the reaction of $i$ to arguments, which will serve as the informational basis to define $\ileadstoall$. Given a proposition $\prop$ and two arguments $\ar, \ar' \in \allargs$, a query consists in presenting both arguments to $i$ and observing which proposition $i$ considers valid in her current state of mind, thus using both arguments and possibly other arguments she has in mind: is it that $\prop$ holds, that $¬\prop$ holds, or that both are possible? The third possibility allows for the case where no argument appears more decisive than the other one.

@@

Define $\ileadstoall$ from $\ileadstoprop$ as follows, for any $\prop \in \allprops$: $(\ar, \ar') \ileadstoall (\prop, \text{possible}) ⇔ (\ar, \ar') \ileadstoprop C$ with $C \subseteq \set{\prop, \text{both}}$; and $(\ar, \ar') \ileadstoall (\prop, \text{sure}) ⇔ (\ar, \ar') \ileadstoprop \set{\prop}$.

@@

When “both” is not observable, $\ileadstoall$ is defined as: $(\ar, \ar') \ileadstoall (\prop, \text{possible}) ⇔ (\ar, \ar') \ileadstoall (\prop, \text{sure}) ⇔ (\ar, \ar') \ileadstoprop \set{\prop}$. (A1) and (A2) still hold. We say in this case that $\ileadstoall$ does not distinguish sure from possible. Thus, elements $\phi \in \Phi$ may be considered as simply equal to $\prop$ or $¬\prop$, as $(\prop, \text{sure})$ is treated exactly as $(\prop, \text{possible})$.

@@

\begin{remark}
	In fact, the model is tasked with finding a stable point after interrogating $i$. The process of trying counter-arguments $\ar_1, …$ is $q(i)$. When done, the model claims that $\ar_0 + \ar_2 + …$ is decisive. The model also claims that another process of interrogation does not lead to different answers. This is falsifiable.
	When claiming that $(\ar_1, \ar_0) \ileadstoall[i] \phi$, the model really claims that $\nexists p, i' \in q^{-1}(q(i)) \suchthat p, \ar_1, \ar_0 \ileadsto[i'] ¬\phi \lor p, \ar_0, \ar_1 \ileadsto[i'] ¬\phi$, where $i' \in q^{-1}(q(i))$ iff $q(i) = q(i')$.
